---
title: "Motivating Example"
subtitle: "Midwest Big Data Summer School"
author: "Heike Hofmann (hofmann@iastate.edu, @heike_hh), Eric Hare"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    transition: default
    widescreen: true
css:
  styles.css
---

## Motivating Example

- Kick things off by looking at a real data set in R
- Goal: get the flavor of using R for data management and exploration
- Don't worry about understanding all the coding right away
- We will come back to see how everything works in detail

## Tips Dataset

- Tips data set recorded by a server in a restaurant over a span of about 10 weeks
- Server recorded several variables about groups he served:
    - Amount he was tipped
    - Overall bill
    - Several characteristics about the groups being served
- __Primary Question__: How do these variable influence the amount being tipped?

- Follow along (copy & paste the code into the console):
```{r, eval = FALSE}
curl::curl_download(
  "https://raw.githubusercontent.com/heike/rwrks/gh-pages/summerschool/01-Introduction-to-R/code/1-example.R",
  "1-example.R"
)
file.edit("1-example.R")
```


## First look at data in R

Lets use R to look at the top few rows of the tips data set. First, we load the tips data  `read.csv`:

```{r}
tips <- read.csv("http://heike.github.io/rwrks/summerschool/data/tips.csv")
```

Now, we use the `head` function to look at the first 6 rows of the data:

```{r}
head(tips)
```

## Tips Data Attributes

How big is this data set and what types of variables are in each column?

```{r}
str(tips)
```

## Tips Variables

Let's get a summary of the values for each variable in tips:

```{r}
summary(tips)
```

# How does the overall amount affect the tip?

## Scatterplots: tip versus total bill 

First, we need to install and load `ggplot2`:

```{r, eval=FALSE}
install.packages("ggplot2")
library(ggplot2)
```

```{r, echo=FALSE}
library(ggplot2)
```

```{r, fig.height=3, fig.width=7}
qplot(x = total_bill, y = tip, geom = "point", data = tips)
```

## 

```{r, fig.height=3, fig.width=7, echo = FALSE}
qplot(x = total_bill, y = tip, geom = "point", data = tips)
```

There is a surprising amount of scatter!

Correlation is below 0.70
```{r}
cor(tips$total_bill, tips$tip)
```


## Points and Line

Add linear regression line to the plot

```{r, fig.height=4, fig.width=7}
qplot(total_bill, tip, geom = "point", data = tips) + 
    geom_smooth(method = "lm")
```

## Closer Look

```{r, fig.height=5, fig.width=5, echo = FALSE}
qplot(x = total_bill, y = tip, geom = "point", data = tips) +
  scale_y_continuous(breaks=0:10)
```

What do the horizontal lines mean?


## More Scatterplots

Color the points by lunch and dinner groups

```{r, fig.height=4, fig.width=7}
qplot(total_bill, tip, geom = "point", data = tips, colour = time)
```


## Rate of Tipping

Tipping is generally done using a rule of thumb based on a percentage of the total bill. We  create a new variable in the data set for the tipping rate = tip / total bill

```{r}
tips$rate <- tips$tip / tips$total_bill

summary(tips$rate)
```

## Tipping Rate Histogram

Lets look distribution of tipping rate values with a histogram

```{r, fig.height=4, fig.width=7}
qplot(rate, data = tips, binwidth = .01)
```

## Someone is an AMAZING tipper...

One person tipped over 70%, who are they?

```{r}
tips[which.max(tips$rate),]
```

## Gender differences?

Color the points by gender of the person who pays the bill

```{r, fig.height=4, fig.width=7}
qplot(total_bill, tip, geom = "point", data = tips, colour = sex)
```

## Rates by Gender

Look at the average tipping rate for men and women seperately

```{r}
mean(tips$rate[tips$sex == "Male"])
mean(tips$rate[tips$sex == "Female"])
```


## Statistical Significance

There is a difference but is it statistically significant?

```{r}
t.test(rate ~ sex, data = tips)
```

## Multiple co-variates

We can use facetting to find whether rates of tipping depend on gender and smoking
```{r, fig.height=4, fig.width=4}
qplot(total_bill, tip, geom = "point", data = tips, facets=smoker~sex) 
```

## Your Turn {data-background=#527a7a}

Try playing with chunks of code from this session to further investigate the tips data:

1. Get a summary of the total bill values
2. Find the average tip value for smokers
3. Get scatterplots of tip versus total bill  for different days of the week
