---
title: "Your Turn Solutions"
author: "Heike Hofmann, Eric Hare"
date: "June 21, 2016"
output: html_document
---

## Motivating Example

Try playing with chunks of code from this session to further investigate the tips data:

1. Get a summary of the total bill values

```{r}
tips <- read.csv("http://heike.github.io/rwrks/01-r-intro/data/tips.csv")

summary(tips$total_bill)
```

2. Find the average tip value for smokers

```{r}
tips$rate <- tips$tip / tips$total_bill
mean(tips$rate[tips$smoker=="Yes"])
```

3. Get scatterplots of tip versus total bill for different days of the week

```{r}
library(ggplot2)

qplot(total_bill, tip, geom = "point", data = tips, facets=~day)
```

## R Basics

Using the R Reference Card (and the Help pages, if needed), do the following:

1. Find out how many rows and columns the `iris' data set has.  Figure out at least 2 ways to do this.
*Hint: "Variable Information" section on the first page of the reference card!*

```{r}
nrow(iris)
ncol(iris)
dim(iris)
```

2. Use the `rep` function to construct the following vector:  1 1 2 2 3 3 4 4 5 5
*Hint: "Data Creation" section of the reference card*

```{r}
rep(1:5, each = 2)
```

3. Use `rep` to construct this vector: 1 2 3 4 5 1 2 3 4 5 1 2 3 4 5

```{r}
rep(1:5, times = 3)
```

1. Find out how many people tipped over 20%.
*Hint: if you use the `sum` function on a logical vector, it'll return how many TRUEs are in the vector:*
```{r}
sum(c(TRUE, TRUE, FALSE, TRUE, FALSE))
```

```{r}
sum(tips$rate > .2)
```

2. **More Challenging**: Calculate the sum of the total bills of anyone who tipped over 20%

```{r}
sum(tips$total_bill[tips$rate > .2])
```

1. Read up on the dataset (`?diamonds`)
2. Plot price by carat (use qplot - go back to the motivating example for help with the syntax)

```{r}
qplot(carat, price, data = diamonds)
```

3. Create a variable `ppc` for price/carat. Store this variable as a column in the diamonds data

```{r}
diamonds$ppc <- diamonds$price / diamonds$carat
```

4. Make a histogram of all ppc values that exceed $10000 per carat.

```{r}
qplot(ppc, geom = "histogram", data = diamonds[diamonds$ppc > 10000,])
```

5. Explore any other interesting relationships you find

## Data Structures

1. Make a data frame with column 1: 1,2,3,4,5,6 and column 2: a,b,a,b,a,b

```{r}
mydf <- data.frame(col1 = 1:6, col2 = rep(c("a", "b"), times = 3))
```

2. Select only rows with value "a" in column 2 using logical vector

```{r}
mydf[mydf$col2 == "a",]
```

3. `mtcars` is a built in data set like `iris`: Extract the 4th row of the `mtcars` data.

```{r}
mtcars[4,]
```

1. Create a list containing a vector and a 2x3 data frame

```{r}
mylist <- list(vec = 1:6, df = data.frame(x = 1:2, y = 3:4, z = 5:6))
```

2. Use indexing to select the data frame from your list

```{r}
mylist[[2]]
```

3. Use further indexing to select the first row from the data frame in your list

```{r}
mylist[[2]][1,]
```

1. View the top 8 rows of mtcars data

```{r}
head(mtcars, n = 8)
```

2. What type of object is the mtcars data set?

```{r}
str(mtcars)
```

3. How many rows are in the mtcars data set? (try finding this using dim or indexing + length)

```{r}
dim(mtcars)
```

4. Summarize the values in each column in the mtcars data set

```{r}
summary(mtcars)
```

## Packages and Programming

1. Create a function that takes numeric input and provides the mean and a 95% confidence interval for the mean for the data (the t.test function could be useful)

```{r}
mean_and_ci <- function(x) {
    themean <- mean(x)
    theci <- t.test(x)$conf.int
    
    return(list(mean = themean, ci = theci))
}
```

2. Add checks to your function to make sure the data is either numeric or logical.  If it is logical convert it to numeric.

```{r}
mean_and_ci <- function(x) {
    if (!is.numeric(x) && !is.logical(x)) stop("Need logical or numeric data")
    
    x <- as.numeric(x)
    
    themean <- mean(x)
    theci <- t.test(x)$conf.int
    
    return(list(mean = themean, ci = theci))
}
```

3. Loop over the columns of the diamonds data set and apply your function to all of the numeric columns.

```{r}
for (i in 1:ncol(diamonds)) {
    if (is.numeric(diamonds[,i])) print(mean_and_ci(diamonds[,i]))
}
```

## RMarkdown Basics

Study the second page of the [R Markdown Reference Guide](rmarkdown-reference.pdf) and go back to the [Hello R Markdown example](hello_rmarkdown.Rmd)

**Easy**: Modify the figure sizing and alignment.

```{r, fig.width=5, fig.height=3, fig.align='center'}
plot(cars)
```

**Medium**: Add a figure caption.

```{r, fig.cap='This is a caption'}
plot(cars)
```

**Hard**: Can you create an animation? (Hint: look at the `fig.show` chunk option -- you might need to install the **animation** package for this)

