---
title: "Basic Plots in ggplot2"
subtitle: "Midwest Big Data Summer School"
author: "Heike Hofmann (hofmann@iastate.edu, @heike_hh), Eric Hare"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    transition: default
    widescreen: true
css:
  styles.css
---

```{r, echo=FALSE}
library(ggplot2)
```

## Let's Run the Setup File...

You should see a plot appear if setup is successful.

## ggplot2 In a Nutshell

- Wildly popular package for statistical graphics: over 1.75 million downloads from CRAN in 2015 (~ 6,200 times per day)
- Developed by Hadley Wickham (An ISU Alumni)
- Designed to adhere to good graphical practices
- Supports a wide variety plot types
- Constructs plots using the concept of layers\medskip
- http://ggplot2.org/book/ or Hadley's book *ggplot2: Elegant Graphics for Data Analysis* for reference material

## *qplot* Function

The *qplot()* function is the basic workhorse of ggplot2

- Produces all plot types available with ggplot2
- Allows for plotting options within the function statement
- Creates an object that can be saved
- Plot layers can be added to modify plot complexity

## *qplot* Structure

The *qplot()* function has a basic syntax:

*qplot(variables, plot type, dataset, options)*

- variables: list of variables used for the plot
- plot type: specified with a *geom =* statement
- dataset: specified with a *data =* statement
- options: there are so, so many options!

## Diamonds Data

We will explore the diamonds data set (preloaded along with ggplot2) using qplot for basic plotting.

The data set was scraped from a diamond exchange company data base by Hadley.  It contains the prices and attributes of over 50,000 diamonds

## Examining the Diamonds Data

What does the data look like?

Lets look at the top few rows of the diamond data frame to find out!

```{r}
head(diamonds)
```

## Basic Scatterplot

Basic scatter plot of diamond price vs carat weight

```{r, fig.height=4, fig.width=7}
qplot(carat, price, geom = "point", data = diamonds)
```

## Another Scatterplot

Scatter plot of diamond price vs carat weight showing versitility of options in qplot

```{r, fig.height=3, fig.width=7}
qplot(carat, log(price), geom = "point", data = diamonds, 
	alpha = I(0.2), colour = color, 
	main = "Log price by carat weight, grouped by color") + 
    xlab("Carat Weight") + ylab("Log Price")
```

## Your Turn {data-background=#527a7a}

All of the your turns for this section will use the tips data set:

```{r}
tips <- read.csv("http://heike.github.io/rwrks/summerschool/data/tips.csv")
```

1. Use qplot to build a scatterplot of variables tips and total bill
2. Use options within qplot to color points by smokers
3. Clean up axis labels and add main plot title

## Basic Histogram

Basic histogram of price

```{r, fig.height=4, fig.width=7}
qplot(price, geom = "histogram", data = diamonds)
```

## Why setting the bin width is useful

Histogram of price, binwidth is set to $50 

```{r, fig.height=4, fig.width=7}
qplot(price, geom = "histogram",  binwidth = 50, data = diamonds)
```

The gap in prices at around $2000 is due to the scraping procedure.

## Another Histogram

Price histograms faceted by clarity

```{r, fig.height=4, fig.width=7}
qplot(price, geom = "histogram", data = diamonds, binwidth = 100, facets = .~clarity)
```

## Your Turn {data-background=#527a7a}

1. Create a new variable in tips data frame rate = tip / total bill
2. Use qplot to create a histogram of rate
3. Change the bin width on that histogram to 0.05
4. Facet this histogram by size of the group

## Basic Boxplot

Side by side boxplot of diamond prices within clarity groupings

```{r, fig.height=4, fig.width=7}
qplot(clarity, log(price), geom = "boxplot", data = diamonds)
```
Why does price decrease as the quality of the diamonds increases?

## Another Boxplot

Side by side boxplot of log prices within clarity groupings with jittered values overlay

```{r, fig.height=3, fig.width=7}
qplot(clarity, log(price), geom = "boxplot", data = diamonds, 
	main = "Boxplots of log Diamond Prices Grouped by Clarity") +
	geom_jitter(alpha = I(.025))
```
There are two groups of prices ... maybe related to size?

## Another Boxplot

Side by side boxplot of log prices within clarity groupings 

```{r, fig.height=4, fig.width=7}
qplot(clarity, log(price)/carat, geom = "boxplot", data = diamonds)
```



## Your Turn {data-background=#527a7a}

1. Make side by side boxplots of tipping rate for males and females
2. Overlay jittered points for observed values onto this boxplot

## Bar Plots

To investigate bar plots we will switch over to the Titanic data set:

```{r}
titanic <- as.data.frame(Titanic)
```

Data includes passenger characteristics and survival outcomes for those aboard the RMS Titanic's ill fated maiden voyage

## Basic Bar Plot

Basic bar plot of survival outcomes

```{r, fig.height=4, fig.width=7}
qplot(Survived, geom = "bar", data = titanic, weight = Freq)
```

## Another Bar Plot

Bar plot faceted by gender and class

```{r, fig.height=4, fig.width=7}
qplot(Survived, geom = "bar", data = titanic, weight = Freq, 
      facets = Sex~Class)
```

## Your Turn {data-background=#527a7a}

1. Use the tips data to make a barplot for counts of smoking and non smoking customers
2. Facet using day of week and time of day to view how smoking status changes for different meal times
